Measurement of total hadronic differential cross sections in the LArIAT experiment

### 2018-04-11 Chapter-1/1-Theory.tex
s

%%%%%%%%%%
\textbf{CP Violation In Lepton Sector.} The measurement of non-zero value for the oscillation parameter $\theta_{13}$ allows the exploration of low-energy CP violation in the lepton sector at neutrino long baseline oscillation experiments, enabling the possibility to measure the Dirac CP-violating phase $\delta$. Exciting theoretical results tie $\delta$ directly to the generation of the baryon asymmetry of the Universe at the Grand Unified Theory scale \textcolor{red}{a couple of cit would be nice}. According to the theoretical model described in \cite{PASCOLI20071}, for example, leptogenesis can be achieved if $|\sin\theta_{13} \sin \delta| > 0.11$, i.e. $\sin \delta > 0.7$.\\
The asymmetry in the oscillation probability of neutrinos and antineutrinos is the observable sensitive to the Dirac CP-violating phase $\delta$ leveraged in neutrino oscillation experiments. Using the parameterization of the PMNS matrix shown in Equation \ref{eq:PMNS},  the difference in the probability of $\nu_e \rightarrow \nu_\mu$ oscillation and the probability of $\bar\nu_e \rightarrow \bar\nu_\mu$ oscillation can be parametrized as follows \cite{Cervera2000},
\begin{equation}
P_{\nu_e\rightarrow\nu_\mu} - P_{\bar\nu_e\rightarrow\bar\nu_\mu} = J \cos\Big(\pm\delta - \frac{\Delta_{31}L}{2}\Big) \sin\Big(\frac{\Delta_{21}L}{2}\Big)\sin\Big(\frac{\Delta_{31}L }{2}\Big)
\end{equation}
where 
\begin{equation}
J = \cos\theta_{13}\sin2\theta_{13}\sin2\theta_{12}\sin2\theta_{23}
\end{equation}
is the Jarlskog invariant \cite{Jarlskog1985}, $L$ the neutrino baseline and $\Delta_{ab}$ a factor proportional to the sign and magnitude of the mass splitting. 
From these equations, it is clear how the relative  large value of $\theta_{13}$ is a happy accident necessary not to completely suppress the sensitivity to CP violation.  The equations also show how the sensitivity to $\delta$ is tied to the measurement of the least precisely measured mixing angle,  $\theta_{23}$ (via the $\sin2\theta_{23}$ term) and to an other unknown quantity, the neutrino ``mass hierarchy" (via the $\Delta_{ab}$ terms). The precise determination of $\theta_{23}$ is often referred as to ``the octant problem". Current experimental results \textcolor{red}{cite NOVA T2K} are consistent with  $\theta_{23}=45^\circ$, which would imply maximal mixing between $\nu_\mu$ - $\nu_\tau$, hinting to an intriguing new symmetry. Therefore, a precise measurement of $\theta_{23}$  is of great interest for theoretical models of quark-lepton universality [59,84,85,86,87,88] , whose  quark and lepton mixing matrices are proportional to the deviation of $\theta_{23}$ from $45^\circ$.

@@

The SM rather complex group structure, where a gauge group is formed with the direct product of other three groups as shown in eq. \ref{eq:SMGroup},  is unexplained. Also, the SM fails to include a suitable dark matter candidate and a mechanisms that accounts for the baryon asymmetry of the universe. Within the SM, a total of 25 parameters remain seemingly arbitrary and need to be fitted to data: 3 gauge couplings, 9 charged fermion masses, 3 mixing angles and one CP phase in the CKM matrix, the Higgs mass and quartic coupling, $\theta_{QCD}$, 3 neutrino masses, 3 neutrino mixing angles, 1 Dirac phase and, eventually,  2 Majorana phases.

@@

\section{Motivations for Hadronic Cross Sections in Argon}
Critical challenges await  the next decade of experimental physics at the intensity frontier. % will entail an even deeper understanding of the neutrino mixing pattern, investigation of the neutrino mass origin, the determination of the number of neutrinos and their nature, and the measurement of CP violation in the lepton sector. 
%%% Plug to LArTPC %%%%%%%%%%%%
Following the recommendation of the latest Particle Physics Project Prioritization Panel  \cite{P5}, the US  is dedicating substantial resources to the development of a short- and long- baseline neutrino program to address many of open questions in neutrino physics today.  This program pivots on the Liquid Argon Time Projection Chamber (LArTPC) detector technology which will be described in \ref{ch:1}.

@@

Assumptions on the nuclear models and on the interaction of hadrons in the nucleus performed at the level of the neutrino event generator bridge the measurement of the products of a neutrino interaction to the reconstruction of the neutrino energy and flavor. Thus, understanding pion hadronic interactions with the nucleus is particularly important to model correctly resonant, DIS and coherent pion production in neutrino interactions, where the presence of pions in the nucleus is abundant. For example, in case of resonant scattering 
\begin{equation}
\nu_l + N \rightarrow l + \Delta/N^* \rightarrow l+ \pi + N', 
\end{equation}
the $\Delta$ and $N^*$ and excited states will decay hadronically in matters of $\sim10^{-24}$~s inside the nucleus producing pions which will bounce within the nuclear medium. The decay modes for the lower mass $\Delta$ (1232) and $N^*$(1440)  are listed in table \ref{tab:decaysdelta}.
\begin{table}[b]
\centering
\begin{tabular}{|l|l|}
\hline
$\Delta$ (1232) 3/2$^+$                 & $N^*$ (1440) 1/2$^+$    \\ \hline
$\Delta^{++} \text{(uuu)} \rightarrow p \pi^+$ & $N^* \rightarrow N \pi$ \\
$\Delta^{+} \text{(uud)} \rightarrow n \pi^+$  &  $N^* \rightarrow N \pi\pi$\\ 
$\Delta^{+} \text{(uud)} \rightarrow p \pi^0$  &                         \\ 
$\Delta^{0} \text{(udd)} \rightarrow n \pi^0$  &                         \\ 
$\Delta^{0} \text{(udd)} \rightarrow p \pi^-$  &                         \\ 
$\Delta^{-} \text{(ddd)} \rightarrow n \pi^-$  &                         \\ \hline
\end{tabular}
\caption{Main decay modes of the lightest Delta resonance and Nucleon excited state.}
\label{tab:decaysdelta}
\end{table}
The key elements of a neutrino event generators for resonance and DIS events are the nuclear model and the hadron treatment (both production and transportation). As one of the most popular event generators for LAr experiments, we take GENIE~\cite{Andreopoulos:2009rq} here as an example.
The nuclear model used by GENIE for all processes is a relativistic Fermi gas (RFG) modified to incorporate short range nucleon-nucleon correlations \cite{PhysRevD.24.1400}. This means that the initial momentum and binding energy of the struck nucleon is determined by assuming nucleons inside the nucleus are quasi-free, acting independently in the mean field of the nucleus. For A $ >$ 20 like argon, the 2-parameter Woods-Saxon shell model for density function is used.  The GENIE module INTRANUKE \cite{Juszczak2006} is used to simulate the final-state interactions (FSI) which is the hadron re-interaction inside the nucleus. This module places the outgoing particles in the nucleus and propagates them.  In the INRANUKE model, hadrons cannot undergo more than one FSI per event. When possible, external hadron-nucleus scattering data are used to tune INTRANUKE. Since no data is available for Argon, GENIE uses an interpolation of data from heavier and lighter nuclei for the pion-argon cross section leading to big uncertainties in the INTRANUKE module.

>>>>

%%%%%%%%%%
\textbf{CP Violation In Lepton Sector.} The measurement of non-zero value for the oscillation parameter $\theta_{13}$ allows the exploration of low-energy CP violation in the lepton sector at neutrino long baseline oscillation experiments, enabling the possibility to measure the Dirac CP-violating phase $\delta$. Exciting theoretical results tie $\delta$ directly to the generation of the baryon asymmetry of the Universe at the Grand Unified Theory scale \textcolor{red}{a couple of cit would be nice}. According to the theoretical model described in \cite{PASCOLI20071}, for example, leptogenesis can be achieved if $|\sin\theta_{13} \sin \delta| > 0.11$, i.e. $\sin \delta > 0.7$.\\
The asymmetry in the oscillation probability of neutrinos and antineutrinos is the observable sensitive to the Dirac CP-violating phase $\delta$ leveraged in neutrino oscillation experiments. Using the parameterization of the PMNS matrix shown in Equation \ref{eq:PMNS},  the difference in the probability of $\nu_e \rightarrow \nu_\mu$ oscillation and the probability of $\bar\nu_e \rightarrow \bar\nu_\mu$ oscillation can be parametrized as follows \cite{Cervera2000},
\begin{equation}
P_{\nu_e\rightarrow\nu_\mu} - P_{\bar\nu_e\rightarrow\bar\nu_\mu} = J \cos\Big(\pm\delta - \frac{\Delta_{31}L}{2}\Big) \sin\Big(\frac{\Delta_{21}L}{2}\Big)\sin\Big(\frac{\Delta_{31}L }{2}\Big)
\end{equation}
where 
\begin{equation}
J = \cos\theta_{13}\sin2\theta_{13}\sin2\theta_{12}\sin2\theta_{23}
\end{equation}
is the Jarlskog invariant \cite{Jarlskog1985}, $L$ the neutrino baseline and $\Delta_{ab}$ a factor proportional to the sign and magnitude of the mass splitting. 
From these equations, it is clear how the relative  large value of $\theta_{13}$ is a happy accident necessary not to completely suppress the sensitivity to CP violation.  The equations also show how the sensitivity to $\delta$ is tied to the measurement of the least precisely measured mixing angle,  $\theta_{23}$ (via the $\sin2\theta_{23}$ term) and to an other unknown quantity, the neutrino ``mass hierarchy" (via the $\Delta_{ab}$ terms). The precise determination of $\theta_{23}$ is often referred as to ``the octant problem". Current experimental results \textcolor{red}{cite NOVA T2K} are consistent with  $\theta_{23}=45^\circ$, which would imply maximal mixing between $\nu_\mu$ - $\nu_\tau$, hinting to an intriguing new symmetry. Therefore, a precise measurement of $\theta_{23}$  is of great interest for theoretical models of quark-lepton universality \textcolor{red}{change cit [59,84,85,86,87,88]}, whose  quark and lepton mixing matrices are proportional to the deviation of $\theta_{23}$ from $45^\circ$.

@@

The SM does not include a suitable dark matter candidate and a mechanisms that accounts for the baryon asymmetry of the universe.  Additionally, a total of \textcolor{red}{25} parameters remain seemingly arbitrary and need to be fitted to data: 3 gauge couplings, 9 charged fermion masses, 3 mixing angles and one CP phase in the CKM matrix, the Higgs mass and quartic coupling, $\theta_{QCD}$, 3 neutrino masses, 3 neutrino mixing angles, 1 Dirac phase and, eventually,  2 Majorana phases.

From a group theory perspective, the SM has a rather complex group structure, where a gauge group is formed with the direct product of other three groups as shown in eq. \ref{eq:SMGroup}. Drawing a parallel with the electroweak symmetry breaking mechanism, where the $ SU(2)_T \otimes U(1)_Y$ is recovered from $U(1)_{EM}$, an interesting line of simplification for the SM group structure would be to think about a similar mechanism where $SU(3)_C  \otimes SU(2)_T \otimes U(1)_Y$ is recovered from an hypothetical larger group. Just as the electroweak unification becomes evident at energies higher than the Higgs VEV,  a direct manifestation of Grand Unification Theories (GUTs) would occur at even higher energies.

As the smallness of neutrino masses suggests to the existence of a higher mass scale \textcolor{red}{IS THIS TRUE?}, an other, even stronger, hint to Grand Unification comes from the slope of running of the coupling constants. The coupling constants for the electromagnetic, weak and strong interactions in the SM vary as a function of the interaction energy as shown in figure \ref{fig:constants}; they do not exactly meet, but their trend is interesting enough to push theorists to construct theories where perfect unification is achieved with the addition of new particles.

%the running is parametrized as follows
%where $b_i$ are coefficients for each interaction, and  the constants assume the same value $\alpha_{GUT}$ at energy $E=M_{GUT}$

@@

\section{Motivations for Hadronic Cross Sections in Argon}
Critical challenges await  the next decade of experimental physics at the intensity frontier. % will entail an even deeper understanding of the neutrino mixing pattern, investigation of the neutrino mass origin, the determination of the number of neutrinos and their nature, and the measurement of CP violation in the lepton sector. 
%%% Plug to LArTPC %%%%%%%%%%%%
Following the recommendation of the latest Particle Physics Project Prioritization Panel  \cite{P5}, the US  is dedicating substantial resources to the development of a short- and long- baseline neutrino program to address many of open questions in neutrino physics today.  This program pivots on the Liquid Argon Time Projection Chamber (LArTPC) detector technology which will be described in Chapter \ref{ch:2}.

@@

Assumptions on the nuclear models and on the interaction of hadrons inside the nucleus performed at the level of the neutrino event generator bridge the measurement of the products of a neutrino interaction to the reconstruction of the neutrino energy and flavor. Thus, understanding pion hadronic interactions with the nucleus is particularly important to model correctly resonant, DIS and coherent pion production in neutrino interactions, where the presence of pions in the nucleus is abundant. For example, in case of resonant scattering 
\begin{equation}
\nu_l + N \rightarrow l + \Delta/N^* \rightarrow l+ \pi + N', 
\end{equation}
the $\Delta$ and $N^*$ and excited states will decay hadronically in matters of $\sim10^{-24}$~s inside the nucleus producing pions which will bounce within the nuclear medium. The decay modes for the lower mass $\Delta$ (1232) and $N^*$(1440)  are listed in table \ref{tab:decaysdelta}.
\begin{table}[b]
\centering
\begin{tabular}{|l|l|}
\hline
$\Delta$ (1232) 3/2$^+$                 & $N^*$ (1440) 1/2$^+$    \\ \hline
$\Delta^{++} \text{(uuu)} \rightarrow p \pi^+$ & $N^* \rightarrow N \pi$ \\
$\Delta^{+} \text{(uud)} \rightarrow n \pi^+$  &  $N^* \rightarrow N \pi\pi$\\ 
$\Delta^{+} \text{(uud)} \rightarrow p \pi^0$  &                         \\ 
$\Delta^{0} \text{(udd)} \rightarrow n \pi^0$  &                         \\ 
$\Delta^{0} \text{(udd)} \rightarrow p \pi^-$  &                         \\ 
$\Delta^{-} \text{(ddd)} \rightarrow n \pi^-$  &                         \\ \hline
\end{tabular}
\caption{Main decay modes of the lightest Delta resonance and Nucleon excited state.}
\label{tab:decaysdelta}
\end{table}
The key elements of a neutrino event generators for resonance and DIS events are the nuclear model and the hadron treatment (both production and transportation). As one of the most popular event generators for LAr experiments, we take GENIE~\cite{Andreopoulos:2009rq} here as an example.
The nuclear model used by GENIE for all processes is a relativistic Fermi gas (RFG) modified to incorporate short range nucleon-nucleon correlations \cite{PhysRevD.24.1400}. This means that the initial momentum and binding energy of the struck nucleon is determined by assuming nucleons inside the nucleus are quasi-free, acting independently in the mean field of the nucleus. For A $ >$ 20 like argon, the 2-parameter Woods-Saxon shell model for density function is used.  The GENIE module INTRANUKE \cite{Juszczak2006} is used to simulate the final-state interactions (FSI) which is the hadron re-interaction inside the nucleus. This module places the outgoing particles in the nucleus and propagates them.  In the INRANUKE model, hadrons cannot undergo more than one FSI per event. When possible, external hadron-nucleus scattering data are used to tune INTRANUKE. Since no data is available for Argon, GENIE uses an interpolation of data from heavier and lighter nuclei for the pion-argon cross section leading to big uncertainties in the INTRANUKE module.

Once the pion has left the target nucleus, the pion-argon hadronic cross section plays an important role in the pion transportation inside the argon medium: processes like pion absorption with emission of protons or pion charge exchange can greatly modify the topology of a neutrino interactions in the detector and lead to errors in the event classification.
Being able to reconstruct the details of pions inside the detector is an imperative for modern liquid argon neutrino experiment to achieve the design resolution for their key physics measurements.
